---
title: "Compulsory exercise 1: Group 12"
author: "Evgeni Vershinin, Kristina Ødegård"
date: "2024-02-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Problem 1  
*a)  
Quantative: Time, income earned, horsepower of a car.
Qvalitative: Marital status, origin, Gender.  
*b) KNN, LDA, QDA can be used for multi-class classifications.  
*c)  
*d)The nereast neighbour for k=1 is a blue dot, so our classification is blue. For K=3, two of the nearest neighbors are red and 1 blue. This gives 2/3 red, so it is red. For K=5 we have 3/5 red, so it is red.


```{r setup2, echo=TRUE, eval=TRUE}
library(MASS)
data(Boston)
data = Boston

model = lm(medv ~ rm + age,data=data)
summary(model)

cor_matrix = cor(data.frame(data$medv, data$rm, data$age))
print(cor_matrix)
model2 = lm(medv ~ rm + age + nox, data=data)
summary(model2)
library(GGally)
ggpairs(data.frame(data$medv, data$rm, data$age, data$nox))
```

**1
e)
IV
Looking at the correlation between Age and NOX, it is 0.731, which is quite high which suggest it has  multicollinearity, which means they give the some of the same information for the model.
